Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads

### 2025-05-01 analysis/paper/paper.qmd
GAM models nearly done; hand edit PCA plot

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Computational reproducibility is increasingly essential for advancing archaeological science, yet its implementation remains inconsistent across the discipline. This study presents a bibliometric assessment of archaeology’s status relative to other fields alongside a systematic review of computational reproducibility in published archaeological research. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

items_df <- 
  read_rds(here::here("analysis/data/wos-data-df.rds"))

@@

source(here::here("analysis/code/008-redraw-Fanelli-and-Glanzel-Fig-2.R"))

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green) Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. 
#| fig-height: 4

library(ggpmisc)
library(mgcv)
formula <-  y ~ x

over_time_long_colour_gams <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(mod_gam = lapply(data, 
                          function(df) gam(value ~ s(year, bs = "cr"), 
                                           data = df)))

over_time_long_colour_gams_summary <- 
over_time_long_colour %>% 
  nest(.by = variable) %>% 
  mutate(fit = map(data, ~mgcv::gam(value ~ s(year, bs = "cs"), data = .)),
         results = map(fit, glance),
         R.square = map_dbl(fit, ~ summary(.)$r.sq)) %>%
  unnest(results) %>%
  select(-data, -fit) %>% 
  select(variable, adj.r.squared)

over_time_long_colour_gams_summary_df <- 
over_time_long_colour %>% 
  left_join(over_time_long_colour_gams_summary)

ggplot() + 
  geom_point(data = over_time_long_colour_gams_summary_df, 
       aes(year,  
           value, 
           colour = becoming_more_scientific),
       alpha = 0.5) + 
  geom_smooth(data = over_time_long_colour_gams_summary_df,
              aes(year, value),
              method="gam", 
              formula = y ~ s(x, bs = "cs"),
              se = FALSE,  
              size = 2, 
              colour = "#7570b3") +
  facet_wrap( ~ variable,
              scales = "free_y") + 
  theme_bw(base_size = base_size) +
  scale_color_manual(values = c("#d95f02",
                                "#1b9e77" )) +
  guides(colour = "none") +
  ylab("") +
  geom_text(data = over_time_long_colour_gams_summary_df %>% 
              group_by(variable) %>% 
              summarise(max_value = max(value),
                        adj.r.squared = unique(adj.r.squared)),
           aes(
           x = 1980, 
           y = max_value, 
           label = paste("Pseudo R² = ", 
                         signif(adj.r.squared, 
                                digits = 3))),
           hjust = 0, 
           vjust = 1.5,
           size = 2)

@@

The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics. One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010. This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g. *Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists. For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open]. The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in range and diversity of references.

@@

# Plot the PCA results
plot_pca_means <- 
ggplot() + 
  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)")) +
  geom_point(data = pca_scores_means,
               aes(.fittedPC1,
                   .fittedPC2),
             size = 1) +
  geom_text_repel(data = pca_scores_means,
            aes(.fittedPC1,
                .fittedPC2,
                label = journal),
            lineheight = 0.8,
            segment.color = NA,
            size = 1,
            bg.color = "white",  # Color of the halo
            bg.r = 0.2) +
     geom_segment(data = pca_loadings_means, 
               aes(x = 0, 
                   y = 0, 
                   xend = PC1, 
                   yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "grey70") +
  geom_text_repel(data = pca_loadings_means,
            aes(PC1,
                PC2,
                label = column),
            size = 1,
            lineheight = 0.7,
            force = 0,
            segment.color = NA,
            color = "grey40",
           bg.color = "white",  # Color of the halo
          bg.r = 0.2) +
  theme_minimal(base_size = base_size) +
  coord_equal()

@@

```{r}
#| label: fig-variation-by-journal
#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals. The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base. Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm. Panel G is a biplot of the first and second principal components of a PCA computed on the journal means of the five bibliometric variables."
#| fig-height: 6

@@

plot_grid(
  plot_grid(plot_journals_authors, 
           plot_journals_article_length,
           plot_journals_title_length,
           plot_journals_ref_recency, 
           plot_journals_ref_diversity,
           plot_borda_scores,
           nrow = 2,
           labels = LETTERS[1:6],
           label_size = 6),
  plot_pca_means,
  nrow = 2,
  labels = c("", LETTERS[7]),
  rel_heights = c(1, 0.5),
  label_size = 6)

@@

@@

The PCA results in panel G show of @fig-variation-by-journal that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), and more specialised journals in the positive range (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.

@@

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is linked to consensus because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often and more rapidly established. The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

>>>>

---
title: "Is Archaeology a science? Insights and imperatives from 10,000 articles and a year of reproducibility reviews"
author:
  - Ben Marwick:
      correspondence: "yes"
      email: bmarwick@uw.edu
      orcid: 0000-0003-1689-0557
      institute: uw
institute:
  - uw:
      name: Department of Anthropology, University of Washington
      address: 4218 Memorial Way Northeast, Seattle, WA 98105
fig-dpi: 600
title-block-published: "Last updated"  
date: now
date-format: long
format: 
  docx:
    reference-doc: "templates/template.docx" # Insert path for the DOCX file
execute:
  echo: false
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - templates/scholarly-metadata.lua
  - templates/author-info-blocks.lua
  - templates/pagebreak.lua
bibliography: references.bib
csl: "templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  The status of archaeology as a science has been debated for decades. This study presents a bibliometric assessment of archaeology’s status relative to other fields, alongside a systematic review of computational reproducibility in published archaeological research. Reproducibility is a factor in the hardness/softness of a field because of it importance in establishing consensus. Analyzing nearly 10,000 articles, I identify trends in authorship, citation practices, and related metrics that position archaeology between the natural and social sciences. A review of reproducibility assessments for the Journal of Archaeological Science reveals persistent challenges, including missing data, unspecified dependencies, and inadequate documentation. To address these issues, I recommend to authors basic practical steps such as standardized project organization and explicit dependency documentation. Strengthening reproducibility will enhance archaeology’s scientific rigor and ensure the verifiability of research findings. This study underscores the urgent need for cultural and technical shifts to establish reproducibility as a cornerstone of rigorous, accountable, and impactful archaeological science.
keywords: |
  Computational reproducibility; Archaeological science; Bibliometric analysis; Open science
highlights: |
  Archaeology occupies an intermediate position between the natural and social sciences.
  Computational reproducibility in archaeology is low, with key barriers in incomplete compendiums, unspecified dependencies, and errors in code.
  Practical steps like logical folder structures, README files, and dependency management tools can substantially improve computational reproducibility.
  A cultural shift toward valuing reproducibility in hiring, promotion, and training is essential for advancing archaeological science.
---

@@

# these data were prepared using the code in 000-import-raw-data.R
items_df <- 
  read_rds(here::here("analysis/data/wos-data-df.rds"))

@@

source(here::here("analysis/code/001-redraw-Fanelli-and-Glanzel-Fig-2.R"))

@@

```{r}
#| label: fig-change-over-time
#| fig-cap: Distribution of article characteristics for archaeology articles over time. Data points represent individual articles. The colour of the points indicates if the overall trend is toward softer (orange) or harder (green). Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. For recency of references a Zero-One Inflated Beta distribution family was used with a logit link function. For diversity of references and relative title length (ln) the Gaussian family was used with the canonical identity link function. For the number of authors pages the Negative Binomial family was used with a standard log link function. Further details about the model specifications and diagnostics are presented in the supplementary materials. 
#| fig-height: 4

# get this from the supplement

@@

The most striking change over time is in the increase in the number of authors, which has the highest r-squared value of these metrics. One interesting detail evident in @fig-change-over-time is the increase in the range of diversity of references after about 2010. This may be due to some broader changes in academic publishing around this time, such as moves to digital-first continuous publishing, new journals appearing (e.g. *Archaeological and Anthropological Sciences* in 2009 and *Journal of Island & Coastal Archaeology* in 2010), and non-archaeology journals becoming more relevant to archaeologists. For example, *PLOS ONE* received its first impact factor in 2010 and in 2011 Nature's *Scientific Reports* began publishing [@malashichev2017open]. The appearance of Google Scholar in 2004, increasing the discoverability of many works for many researchers, may have also contributed to this increase in diversity of references.

@@

# Plot the PCA results
plot_pca_means <- 
ggplot() + 
  labs(x = paste0("PC1 (", variance_explained[1], "%)"),
       y = paste0("PC2 (", variance_explained[2], "%)")) +
  geom_point(data = pca_scores_means,
               aes(.fittedPC1,
                   .fittedPC2),
             size = 1) +
  geom_text_repel(data = pca_scores_means %>% 
                    mutate(label = str_replace(journal, 
                                               "JOURNAL", 
                                               "J.")) %>% 
                    mutate(label = str_remove(label, 
                                               "-AN\nINTERNATIONAL\nJ.")),
            aes(.fittedPC1,
                .fittedPC2,
                label = label),
            lineheight = 0.8,
            segment.color = NA,
            force_pull = 10,
            size = 2.5,
            bg.color = "white",  # Color of the halo
            bg.r = 0.2) +
     geom_segment(data = pca_loadings_means, 
               aes(x = 0, 
                   y = 0, 
                   xend = PC1, 
                   yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")), 
               color = "grey70") +
  geom_text_repel(data = pca_loadings_means,
            aes(PC1,
                PC2,
                label = column),
            size = 2,
            lineheight = 0.8,
            force = 10,
            force_pull = 0,
            segment.color = NA,
            color = "grey40",
           bg.color = "white",  # Color of the halo
          bg.r = 0.2) +
  theme_minimal(base_size = base_size) +
  coord_fixed(xlim = c(-6, 2.5),
              ylim = c(-3, 2))

# tricky to get the label spacing right, let's save an SVG, edit
# by hand, then export to PNG and read that file later. 
ggsave(plot_pca_means,
       filename = here::here("analysis/figures/plot_pca_means.svg"))

@@

```{r}
#| label: fig-variation-by-journal
#| fig-cap: "Panels A-E: Variation in bibliometric indicators of hardness for 20 archaeological journals. The journals are ordered for each indicator so that within each plot, the harder journals are at the top of the plot and the softer journals are at the base. Panel F shows a bar plot that is the single consensus ranking computed from all five variables, using the Borda Count ranking algorithm."
#| fig-height: 5

@@

  plot_grid(plot_journals_authors, 
           plot_journals_article_length,
           plot_journals_title_length,
           plot_journals_ref_recency, 
           plot_journals_ref_diversity,
           plot_borda_scores,
           nrow = 2,
           labels = LETTERS[1:6],
           label_size = 6)

@@

```{r}
#| label: fig-pca
#| fig-cap: "Biplot of the first and second principal components of a PCA computed on the means of the five bibliometric variables for each journal in the sample.  The arrows represent the correlation between each original variable and the principal components. The direction and length of the arrows indicate how strongly each variable contributes to each component. "
#| fig-height: 4

knitr::include_graphics(here::here("analysis/figures/plot_pca_means.png"))
```

@@

The PCA results in @fig-pca shows that PC1 captures most of the variance in the metrics (`r variance_explained[1]`%) and is reasonable proxy for the hard-soft spectrum, with *Journal of Cultural Heritage* representing the hard extreme on the right and *Journal of Archaeological Research* representing the soft extreme on the left. PC1 is explained by title length, number of pages, and the diversity of references. Journals with higher PC1 values have articles with longer titles, fewer pages, and less diverse reference lists. The distribution of PC1 values is skewed left, with most of the journals concentrated at the harder end of the spectrum. PC2 is explained by number of authors and recency of references. The distribution of PC2 values reveals additional structure to the data and can be roughly separated into generalist journals in the negative range (e.g. *American Antiquity*, *Antiquity*, *Advances in Archaeological Practice*), characterised by fewer authors and more recent references. In the positive range of PC2 are more specialised journals, characterised by higher numbers of authors and less recent references cited (e.g. *Environmental Archaeology*, *Geoarchaeology*, *Archaeologial Research in Asia*, *Journal of Island and Coastal Archaeology*). The *Journal of Archaeological Science* sits about midway between these two groups, reflecting its relevance to specialised communities of practice in archaeology, as well as generalist communities of practice.

@@

While these bibliometric variables provide several interesting insights into the status of archaeology as a science, via measurement of consensus, and are important for moving the debate beyond discussions of a small number of case studies, they miss a crucial factor that separates scientific practice from non-science. This is reproducibility, which, according to a report by the US National Science Foundation [@cacioppo2015social], "refers to the ability of a researcher to duplicate the results of a prior study using the same materials as were used by the original investigator. That is, a second researcher might use the same raw data to build the same analysis files and implement the same statistical analysis in an attempt to yield the same results... Reproducibility is a minimum necessary condition for a finding to be believable and informative." Scientific reproducibility is a factor contributing to the hardness of a field. Specifically, reproducibility is linked to to concept of consensus in a field because if more researchers provide sufficient detail for others to reproduce their results, then consensus on new knowledge can more often, and more, rapidly be established. The importance of this factor can be traced to Irish chemist Robert Boyle (1627-1691), best known for his experiments with vacuum pumps [@shapin2011leviathan]. Boyle was concerned about the secrecy common among experimentalists in the 17th century and aimed to shift the culture from valuing direct in-person witnessing of scientific demonstrations towards meticulous written communications that were detailed enough to enable a reader to successfully undertake the experiment themselves, independent of the original author.

### 2025-05-01 analysis/paper/supplement-GAMS-details.qmd
GAM models nearly done; hand edit PCA plot

---
title: "Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads"
author: "Ben Marwick"
date: "`r Sys.Date()`"
bibliography: ../analysis/paper/references.bib
csl:          ../analysis/paper/templates/journal-of-archaeological-science.csl # Insert path for the bib-style
output:
  html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
```

@@

Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data. @fig-brms-distros shows the distributions of the five variables related to scientific publications. The shape of these distributions informed our decisions about model specification. 
Recency of references: This variable, bounded between 0 and 1, exhibited a right-skewed distribution with notable occurrences at the boundaries (0 and 1). Consequently, a Zero-One Inflated Beta (zero_one_inflated_beta()) distribution family was selected. The associated logit link function is standard for modeling the mean (mu) of proportions, transforming the (0, 1) interval to the real line (-∞, +∞) and modeling the log-odds of the mean recency score for values strictly between 0 and 1.

@@

library(brms)
library(dplyr)
library(purrr)

@@

kable(
  diagnostics_table,
  caption = "Summary of MCMC Convergence Diagnostics. Max R-hat should be close to 1.0. Min ESS values indicate the minimum effective sample size across all parameters. N Divergences should be 0.",
  digits = c(NA, 0, 3, 0, 0, 0), # Chains/ESS/Div as integers, Rhat to 3 decimals
  booktabs = TRUE,
  linesep = ""
) %>%
  kableExtra::kable_styling(latex_options = "scale_down", full_width = FALSE) %>%
  kableExtra::footnote(general = "Ideal values: Max R-hat ≈ 1.0, Min ESS > 1000, N Divergences = 0.", general_title = "Note:", footnote_as_chunk = TRUE)
```

@@

```{r fig.height=3 * length(results_brms_valid$mod_brms)}
#| label: fig-trace-plots-mcmc-trace
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model. Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)     # For map2
library(dplyr)     # For tibble etc.
library(posterior) # For as_draws_array
library(bayesplot) # For mcmc_trace and nuts_params
library(ggplot2)   # For ggtitle and theme adjustments
library(cowplot)   # For plot_grid
#---------------------------------------------------------------------

@@

    # Return the ggplot object (or NULL if failed or no pars found)
    return(p)
  }
)
#---------------------------------------------------------------------

#---------------------------------------------------------------------
# Combine and Display Plots
#---------------------------------------------------------------------
# Remove any NULL elements from the list
valid_trace_plots <- trace_plots_list[!sapply(trace_plots_list, is.null)]

@@

  # Display the combined plot
  print(combined_trace_plots)
} else {
  message("No valid trace plots were generated.")
}
#---------------------------------------------------------------------
```

@@

```{r}
#| label: tbl-param-summary
#| echo: false
#| message: false
#| warning: false
#---------------------------------------------------------------------
# Load necessary libraries
#---------------------------------------------------------------------
library(brms)
library(purrr)
library(dplyr)
library(tibble) # For rownames_to_column
library(knitr)
library(kableExtra)
#---------------------------------------------------------------------

@@

@@

The estimated non-linear effect of publication year on each response variable is visualized below (@fig-smooth-plots), showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale).

```{r}
#| results: "hide"

# Load necessary libraries
library(brms)
library(purrr)
library(dplyr)
library(ggplot2)
library(cowplot)

@@

    # Generate the plot object list directly
    plot_object_list <- plot(conditional_smooths(model), plot = TRUE)

@@

    # Extract the first plot, add title, label, and theme
    plot_object_list[[1]] +
      ggtitle(var_name) +
      ylab(y_axis_label) +
      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
  }
)
```

```{r}
#| label: fig-smooth-plots
#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs. The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval. Y-axis scales differ and correspond to the link function of each model (identity, log, or logit)."

# Combine the plots into a grid (adjust ncol as desired)
# The plot_grid call itself will display the combined plot
plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)

>>>>

---
title: "Supplementary Materials on Bayesian Generalized Additive Models used to Analyse Temporal Trends in Job Ads"
author: "Ben Marwick"
date: "`r Sys.Date()`"
bibliography: references.bib
csl:          templates/journal-of-archaeological-science.csl # Insert path for the bib-style
output:
  html_document 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = 'center')
library(brms)
library(tidyverse)
library(broom.mixed) # Provides tidy methods for brmsfit objects
library(cowplot)     # For plot_grid
library(knitr)       # For kable
library(kableExtra)
library(posterior) # For as_draws_array
library(bayesplot) # For mcmc_trace and nuts_params
```

@@

Separate GAMs were fitted for each response variable against publication year (`year`). The specific model structure for each variable was chosen based on the nature of the response data. @fig-brms-distros shows the distributions of the five variables related to scientific publications. The shape of these distributions informed our decisions about model specification.

Recency of references: This variable, bounded between 0 and 1, exhibited a right-skewed distribution with notable occurrences at the boundaries (0 and 1). Consequently, a Zero-One Inflated Beta (zero_one_inflated_beta()) distribution family was selected. The associated logit link function is standard for modeling the mean (mu) of proportions, transforming the (0, 1) interval to the real line (-∞, +∞) and modeling the log-odds of the mean recency score for values strictly between 0 and 1.

@@

@@

kable(
  diagnostics_table,
  digits = c(NA, 0, 3, 0, 0, 0), # Chains/ESS/Div as integers, Rhat to 3 decimals
  booktabs = TRUE,
  linesep = ""
) %>%
  kableExtra::kable_styling(latex_options = "scale_down", full_width = FALSE) %>%
  kableExtra::footnote(general = "Ideal values: Max R-hat ≈ 1.0, Min ESS > 1000, N Divergences = 0.", general_title = "Note:", footnote_as_chunk = TRUE)
```

@@

```{r fig.height=3 * length(results_brms_valid$mod_brms)}
#| label: fig-trace-plots-mcmc-trace
#| echo: false
#| message: false
#| warning: false
#| fig-width: 10
#| fig-cap: "Trace plots for key parameters (Intercept and smooth term standard deviation) with divergence markers for each model. Well-mixing chains should resemble stationary 'fuzzy caterpillars' with chains overlapping."

@@

    # Return the ggplot object (or NULL if failed or no pars found)
    return(p)
  }
)

# Combine and Display Plots
#---------------------------------------------------------------------
# Remove any NULL elements from the list
valid_trace_plots <- trace_plots_list[!sapply(trace_plots_list, is.null)]

@@

  # Display the combined plot
  combined_trace_plots
} else {
  message("No valid trace plots were generated.")
}
#---------------------------------------------------------------------
```

@@

```{r}
#| label: tbl-param-summary
#| message: false
#| warning: false

@@

# R-squared Equivalent for Model Fit

Standard R-squared isn't directly applicable to many GAM families (especially non-Gaussian ones). The recommended metric for brms models is the Bayesian R-squared. Calculated using brms::bayes_R2(model), this metric represents the variance of the predicted values divided by the sum of the variance of the predicted values and the expected variance of the residuals. It estimates the proportion of outcome variance that can be explained by the model's predictors. The results are shown below in @tbl-r-sq.

```{r}
#| label: tbl-r-sq
#| tbl-cap: "Bayesian R-squared estimates for each model."
#| message: false
#| warning: false

# Calculate Bayesian R-squared for each model
# Use map to apply to each model, map_dfr to combine
bayes_r2_summary <- map2_dfr(
  results_brms_valid$mod_brms,
  results_brms_valid$variable,
  ~{
    model <- .x
    var_name <- .y
    r2_est <- NA_real_
    r2_lower <- NA_real_
    r2_upper <- NA_real_

           # Calculate Bayesian R2 (returns matrix with Estimate, Est.Error, Q2.5, Q97.5)
           r2_obj <- brms::bayes_R2(model, ndraws = 10000)
           # Extract the point estimate and CI bounds
           r2_est <- r2_obj[1, "Estimate"]
           r2_lower <- r2_obj[1, "Q2.5"]
           r2_upper <- r2_obj[1, "Q97.5"]

    tibble(
        Variable = var_name,
        `Bayes R2` = r2_est,
        `Lower 95% CI` = r2_lower,
        `Upper 95% CI` = r2_upper
        )
  }
)

kable(bayes_r2_summary, 
      digits = 2, 
      booktabs = TRUE) %>%
kable_styling(latex_options = "scale_down")
```

@@

The estimated non-linear effect of publication year on each response variable is visualized below (@fig-smooth-plots), showing the contribution of the smooth term `s(year)` to the model's linear predictor (on the respective link scale). This allows us to visually assess the significance of the model.

Recency of references: There is strong evidence for a significant temporal trend. The 95% CI falls convincingly below zero from approximately the early 1990s onwards, indicating a significant decrease in the contribution of year to recency (on the logit scale) during the later decades of the study period. Evidence for an effect differing from zero in the earliest years is less conclusive as the CI approaches the zero line.

Diversity of references: The effect of year on reference diversity shows strong evidence of being non-zero across most of the timeframe. The 95% CI is clearly separated from zero, initially below and later rising significantly above it, particularly after 2010, indicating substantial temporal variation not attributable to chance.

Relative title length (ln): The evidence suggests a significant effect of year primarily concentrated around 1990, where the 95% CI lies entirely below the zero line. During the initial years and from roughly 2000 onwards, the CI overlaps zero, indicating that the evidence for an effect of year being different from zero is weak or uncertain in these periods.

N. of authors: Strong evidence supports a significant positive effect of year on the number of authors (on the log scale). The 95% CI is consistently above the zero line from the late 1980s onwards, demonstrating a significant increasing trend over the majority of the period examined.

N. of pages: The temporal effect on the number of pages appears significant during specific periods. The 95% CI is clearly above zero during the peak in the 1990s, providing strong evidence for an increased effect during that time. However, the CI overlaps zero during the trough in the mid-2000s and potentially near the end of the period, suggesting uncertainty about whether the effect differs significantly from zero during these phases. There is also evidence for a negative effect relative to the intercept in the earliest years.

```{r}
#| label: fig-smooth-plots
#| fig-cap: "Estimated conditional effect of year (s(year)) on each response variable from the Bayesian GAMs. The blue line represents the posterior mean effect, and the grey ribbon represents the 95% credible interval. Y-axis scales differ and correspond to the link function of each model (identity, log, or logit). The red line is y=0 to visually assess the significance of the  (i.e., whether the credible interval excludes zero)."

@@

    # Generate the plot object list directly
    plot_object_list <- plot(conditional_smooths(model), plot = FALSE)

@@

    # Extract the first plot, add title, label, and theme
    plot_object_list[[1]] + 
      geom_hline(yintercept = 0, 
                 color = "red", 
                 linetype = "dashed", 
                 linewidth = 0.8) +
      ggtitle(var_name) +
      ylab(y_axis_label) +
      theme_minimal() +
      theme(plot.title = element_text(hjust = 0.5, size = 10)) # Center title
  }
)

# plot a panel of all the plots
plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
```

```{r}
# plot for manuscript
# Generate conditional smooth plots within pmap

input_list <-   list(
    model = results_brms_valid$mod_brms,
    var_name = results_brms_valid$variable,
    raw_data = results_brms_valid$data,
    r_sq = bayes_r2_summary$`Bayes R2`
  )

conditional_smooths_plots_fn <- function(model, var_name, raw_data, r_sq){

# Generate the plot object list 
     smooth_data <- conditional_smooths(model, plot = FALSE)[[1]]
     intercept_est <- fixef(model)["Intercept", "Estimate"]

# --- Determine inverse link function and response label ---

        inv_link <- identity # Default for Gaussian
        response_label <- var_name # Default label

        if (var_name %in% c("N. of authors", "N. of pages")) {
          inv_link <- exp # Inverse of log link
          response_label <- paste("Estimated", var_name)
        } else if (var_name == "Recency of references") {
          inv_link <- plogis # Inverse of logit link
          response_label <- "Estimated Recency of references (Price Index)"
        } else if (var_name == "Diversity of references") {
           response_label <- "Estimated Diversity of references"
        } else if (var_name == "Relative title length (ln)") {
           response_label <- "Estimated Rel. Title Length (ln)"
        }

# --- Add intercept and apply inverse link function so we can plot on the scale of the raw data
        smooth_data <- 
          smooth_data %>%
          mutate(
            # Calculate estimates on response scale
            response_est = inv_link(intercept_est + estimate__),
            response_lower = inv_link(intercept_est + lower__),
            response_upper = inv_link(intercept_est + upper__)
          )

    ggplot(smooth_data, 
           aes(x = year, 
               y = response_est)) +
            geom_point(data = raw_data %>% 
                         mutate(colour_hex = case_when(
                           becoming_more_scientific == "TRUE" ~ "#1b9e77",
                           .default = "#d95f02"
                         )),
                 aes(x = year,
                     y = value,
                     colour = colour_hex),
                 alpha = 0.7) +
      scale_color_identity() +
      # Add the line 
      geom_line(color = "#7570b3", 
                linewidth = 2) +
      geom_text(data = raw_data  %>% 
                  mutate(`Bayes R2` = r_sq) %>% 
                  mutate(max_value = max(value)) %>% 
                  distinct(`Bayes R2`, 
                           max_value),
           aes(
           x = 1980, 
           y = max_value, 
           label = paste("Pseudo R² = ", 
                         signif(`Bayes R2`, 
                                digits = 3))),
           hjust = 0, 
           vjust = 1.5,
           size = 2) +
    theme_bw(base_size = 6) +
      labs( 
           y = response_label, 
           x = "")
}

conditional_smooths_plots_list <- 
  pmap(input_list,
       conditional_smooths_plots_fn)
```

Here is the figure that appears in the main manuscript:

```{r}
#| label: fig-smooth-plots-paper
#| fig-cap: "Distribution of article characteristics for archaeology articles over time and estimated conditional effect of year (purple lines). Data points represent individual articles. The colour of the points indicates if the overall temporal trend is toward archaeology become a softer (orange) or harder (green) field. Bayesian Generalized Additive Models were computed to fit the lines summarising the relationships between the variables and the time series. For recency of references, a Zero-One Inflated Beta distribution family was used with a logit link function. For diversity of references and relative title length (ln), the Gaussian family was used with the canonical identity link function. For the number of authors pages the Negative Binomial family was used with a standard log link function. Further details about the model specifications and diagnostics are presented in the supplementary materials."

plot_grid(plotlist = conditional_smooths_plots_list, ncol = 3)
